---
title: "Mavrothalassitis' analysis"
author: "Ismini Baltsavia"
date: "July 18, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Dimensions of each file:
```{r, echo=FALSE}

setwd("/home/ismini/Documents/bioinfo-grad/mavrothalassitis/")
#read files
frog = read.csv("Frog_Clusters.csv", header=T, sep=',')
dim_frog = dim(frog)
cat("dimensions of frog file:") 
dim_frog

#1434 19

#GeneLists_OldProt: read the 6 first rows as header
header1 <- sapply(read.csv("GeneLists_OldProt.csv", header=F, sep=",", 
                          check.names=FALSE, nrow=6) , paste, collapse="_")

genes_OldProt <- read.csv("GeneLists_OldProt.csv", sep=",", header = F,
                     check.names=FALSE, skip=6, col.names=header1)
dim_genes_OldProt = dim(genes_OldProt)
cat("dimensions of GeneLists_OldProt file:") 
dim_genes_OldProt
#140 6

#Gene_lists_quant_batch: read the 4 first rows as header
header2 <- sapply(read.csv("Gene_lists_quant_batch.csv", header=F, sep=",", 
                          check.names=FALSE, nrow=4) , paste, collapse="_")

genes_batch <- read.csv("Gene_lists_quant_batch.csv", sep=",", header = F,
                     check.names=FALSE, skip=4, col.names=header2)
dim_genes_batch = dim(genes_batch)
cat("dimensions of Gene_lists_quant_batch file:") 
dim_genes_batch
#581 32

#GeneLists_oldVSall: read the 6 first rows as header
header3 <- sapply(read.csv("GeneLists_oldVSall.csv", header=F, sep=",", 
                          check.names=FALSE, nrow=6) , paste, collapse="_")

genes_oldVSall <- read.csv("GeneLists_oldVSall.csv", sep=",", header = F,
                     check.names=FALSE, skip=6, col.names=header3)
dim_genes_oldVSall = dim(genes_oldVSall)
cat("dimensions of GeneLists_oldVSall file:") 
dim_genes_oldVSall
#639 34
```

## Fisher's exact test between frog and each of the remaining matrices, p values, adjusted p values and q values
```{r, echo=FALSE}
fun <- function(df1, df2){
  l <- c()
  g<- list()
  c = 0
  for(i in 1:ncol(df1)){
    for (k in 1:ncol(df2)){
      v1 <- as.matrix(df1)[,i]
      v2 <- as.matrix(df2)[,k]
      
      
      common_genes = intersect(v1, v2)
      
      count_g = length(common_genes)
      l<- c(l, count_g)
      
      c = c + 1
      g[c]<- list(common_genes)
      
      #print(count_g)
      #common_genes
    
    }
  }
  l
  #common_genes
  #g
}

count_genes_targets_genes_batch = fun(frog, genes_batch)
count_genes_targets_genes_OldProt = fun(frog, genes_OldProt)
count_genes_targets_genes_OldVSall = fun(frog, genes_oldVSall)


###list of common genes
fun2 <- function(df1, df2){
  g<- list()
  c = 0
  for(i in 1:ncol(df1)){
    for (k in 1:ncol(df2)){
      v1 <- as.matrix(df1)[,i]
      v2 <- as.matrix(df2)[,k]
    
      common_genes = intersect(v1, v2)
      c = c + 1
      g[c]<- list(common_genes)

    }
  }
  g
}

common_genes_targets_genes_batch <- fun2(frog, genes_batch)
common_genes_targets_genes_OldProt = fun2(frog, genes_OldProt)
common_genes_targets_genes_OldVSall = fun2(frog, genes_oldVSall)

#####Fischer's test

my_fishers_test <- function(df1, df2){
  f <- c()
  for(i in 1:ncol(df1)){
    for (k in 1:ncol(df2)){
      v1 <- as.matrix(df1)[,i]
      v2 <- as.matrix(df2)[,k]
      
      n=20465
      f_test <- fisher.test(matrix(c(n-length(union(v1,v2)),   length(setdiff(v1,v2)), 
                length(setdiff(v2,v1)), length(intersect(v1,v2))), nrow=2), 
                alternative="greater")$p.value

      f<- c(f, f_test)
      
      #print(count_g)
      #common_genes
      
    }
  }
  f
}

fishers_targets_genes_batch <- my_fishers_test(frog, genes_batch)
fishers_targets_genes_OldProt = my_fishers_test(frog, genes_OldProt)
fishers_targets_genes_OldVSall = my_fishers_test(frog, genes_oldVSall)
```


```{r, echo=FALSE}
#p values in total
pvalues=c(fishers_targets_genes_batch,fishers_targets_genes_OldProt,fishers_targets_genes_OldVSall)
p_0.05 = length(pvalues[pvalues<0.05])

cat("comparisons with a p value < 0.05:", p_0.05)
#10

hist(pvalues, main="Density plot of p values",breaks=seq(0,1,0.05), freq=FALSE)
```

### Bonferroni adjustment of p values
```{r, echo=FALSE}
#p values adjustment with Bonferroni method
pvalues_bonf <- p.adjust(pvalues,method="bonferroni")
p_0.05_bonf = length(pvalues_bonf[pvalues_bonf<0.05])
cat("comparisons with an adjusted (Bonferroni method) p value < 0.05:", p_0.05_bonf)

hist(pvalues_bonf, main="Density plot of adjusted (Bonferroni method) p values",breaks=seq(0,1,0.05), freq=FALSE)
mean_bonf = mean(pvalues_bonf)
cat("mean of adjusted (Bonferroni method) p values:", mean_bonf)
```

### Benjamini-Hochberg adjustment of p values
```{r, echo=FALSE}
#p values adjustment with Benjamini Hochberg method
pvalues_fdr <- p.adjust(pvalues,method="fdr")
p_0.05_fdr = length(pvalues_fdr[pvalues_fdr<0.05])
cat("comparisons with an adjusted (Benjamini_Hochberg method) p value < 0.05:", p_0.05_fdr)

hist(pvalues_fdr, main="Density plot of adjusted (Benjamini-Hochberg method) p values",breaks=seq(0,1,0.05), freq=FALSE )
mean_fdr = mean(pvalues_fdr)
cat("mean of adjusted (Benjamini-Hochberg method) p values:", mean_fdr)
```

### q values
```{r, echo=FALSE}
#install.packages("qvalue")
#if (!requireNamespace("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")

#BiocManager::install("qvalue")
library(qvalue)
qobj <-qvalue(pvalues)
q_0.05 = length(qobj$qvalues[qobj$qvalues<0.05])
cat("comparisons with a q value < 0.05:", q_0.05)

#hist(qobj)
hist(qobj$qvalues, main="Density plot of q values", breaks=seq(0,1,0.05), freq=FALSE)
mean_qval = mean(qobj$qvalues)
cat("mean of q values:", mean_qval)
```

## References

HarvardX Biomedical Data Science Open Online Training (http://rafalab.github.io/)
https://genomicsclass.github.io/book/pages/multiple_testing.html

Pawitan Y, Michiels S, Koscielny S, Gusnanto A, Ploner A. 2005. False discovery rate, sensitivity and sample size for microarray studies. Bioinformatics, Volume 21: 3017–3024, https://doi.org/10.1093/bioinformatics/bti448

Storey JD, Tibshirani R. 2003. Statistical significance for genomewide studies. Proceedings of the National Academy of Sciences, 100:9440-9445, https://doi.org/10.1073/pnas.1530509100

Storey JD. 2010. False Discovery Rates. Princeton University, Princeton, USA http://genomine.org/papers/Storey_FDR_2011.pdf

Storey JD, Bass AJ, Dabney A, Robinson D. 2019. qvalue: Q-value estimation for false discovery rate control. R package version 2.14.1. http://github.com/jdstorey/qvalue


```{r, eval=TRUE, echo=FALSE}
#mean(qobj$lfdr)
#max(qobj$qvalues[qobj$pvalues <= 0.05])

setwd("/home/ismini/Documents/bioinfo-grad/mavrothalassitis/Analysis_all_datasets/")

###frog vs genes_batch#####################################
max_genes_batch_comparisons <- max(count_genes_targets_genes_batch) #it is 71
common1 = sapply (common_genes_targets_genes_batch, function (x) {length (x) <- max_genes_batch_comparisons; return (x)})

#labels for columnwise comparisons
labels1= c()
for(i in 1:ncol(frog)){
  for (k in 1:ncol(genes_batch)){
    label = paste(colnames(frog)[i], " versus ", colnames(genes_batch)[k])
    labels1 = c(labels1, label)
    labels1
  }
}

#labels, pvalues and common genes in one matrix
l_genes_batch_comparisons = dim_frog[2]*dim_genes_batch[2]
final1 = rbind(labels1, fishers_targets_genes_batch, pvalues_bonf[1:l_genes_batch_comparisons], pvalues_fdr[1:l_genes_batch_comparisons], qobj$qvalues[1:l_genes_batch_comparisons], common1)

#final1 <- rbind(matrix(data=unlist(strsplit(final1[1,], " v ")), nrow=2,ncol=320), final1[-1,])
#split header back to 4 rows 
DF1 <- read.table(text=final1[1,], sep ="_", as.is = TRUE)
final_frog_genes_batch = rbind(t(DF1),  final1[-1,])
write.table(final_frog_genes_batch, "final_frog_genes_batch.csv", sep=",", na="")
```

```{r, echo=FALSE}
###frog vs genes_OldProt##########################

max_genes_OldProt_comparisons <- max(count_genes_targets_genes_OldProt) #it is 71
common2 = sapply (common_genes_targets_genes_OldProt, function (x) {length (x) <- max_genes_OldProt_comparisons; return (x)})

#labels for columnwise comparisons
labels2= c()
for(i in 1:ncol(frog)){
  for (k in 1:ncol(genes_OldProt)){
    label = paste(colnames(frog)[i], " versus ", colnames(genes_OldProt)[k])
    labels2 = c(labels2, label)
    labels2
  }
}

#labels, pvalues and common genes in one matrix
l_genes_OldProt_comparisons = dim_frog[2]*dim_genes_OldProt[2]

final2 = rbind(labels2, fishers_targets_genes_OldProt,      pvalues_bonf[(l_genes_batch_comparisons+1):(l_genes_batch_comparisons+l_genes_OldProt_comparisons)], 
pvalues_fdr[(l_genes_batch_comparisons+1):(l_genes_batch_comparisons+l_genes_OldProt_comparisons)], 
qobj$qvalues[(l_genes_batch_comparisons+1):(l_genes_batch_comparisons+l_genes_OldProt_comparisons)], 
common2)

#split header back to 6 rows 
DF2 <- read.table(text=final2[1,], sep = "_", as.is = TRUE)

final_frog_genes_OldProt = rbind(t(DF2), final2[-1,])
write.table(final_frog_genes_OldProt, "final_frog_genes_OldProt.csv", sep=",", na="")
```


```{r, echo=FALSE}
###frog vs genes_OldVSall################################

max_genes_OldVSall_comparisons <- max(count_genes_targets_genes_OldVSall)
common3 = sapply (common_genes_targets_genes_OldVSall, function (x) {length (x) <- max_genes_OldVSall_comparisons; return (x)})

#labels for columnwise comparisons
labels3= c()
for(i in 1:ncol(frog)){
  for (k in 1:ncol(genes_oldVSall)){
    label = paste(colnames(frog)[i], " versus ", colnames(genes_oldVSall)[k])
    labels3 = c(labels3, label)
    labels3
  }
}

#labels, pvalues and common genes in one matrix
l_genes_OldVSall_comparisons <- dim_frog[2]*dim_genes_oldVSall[2]
final3 = rbind(labels3, 
               fishers_targets_genes_OldVSall, 
               pvalues_bonf[(l_genes_batch_comparisons+l_genes_OldProt_comparisons+1):length(pvalues)], 
               pvalues_fdr[(l_genes_batch_comparisons+l_genes_OldProt_comparisons+1):length(pvalues)], 
               qobj$qvalues[(l_genes_batch_comparisons+l_genes_OldProt_comparisons+1):length(pvalues)], 
               common3)
#split header back to 6 rows 
DF3 <- read.table(text=final3[1,], sep = "_", as.is = TRUE)

final_frog_genes_OldVSall = rbind(t(DF3), final3[-1,])
write.table(final_frog_genes_OldVSall, "final_frog_genes_OldVSall.csv", sep=",", na="")
```

Comparisons with statistical significant p values:
```{r, echo=FALSE}

all_labels <- c(labels1, labels2, labels3)
max_rows <- max(max_genes_batch_comparisons, max_genes_OldProt_comparisons, max_genes_OldVSall_comparisons)
common1 <- apply (common1, 2, function (x) {length (x) <- max_rows; return (x)})
common2 <- apply (common2, 2, function (x) {length (x) <- max_rows; return (x)})
common3 <- apply (common3, 2, function (x) {length (x) <- max_rows; return (x)})

all_common <- cbind(common1, common2, common3)
dim(common3)
sign_p_values_index <- which(pvalues<0.05)

pvalues_mat <- matrix(pvalues, nrow=1, dimnames=(list("1", c(rep("genes_batch",l_genes_batch_comparisons), rep("OldProt", l_genes_OldProt_comparisons), rep("OldVSall", l_genes_OldVSall_comparisons) )) ))

pvalues_bonf_mat <- matrix(pvalues_bonf, nrow=1, dimnames=(list("1", c(rep("genes_batch",l_genes_batch_comparisons), rep("OldProt", l_genes_OldProt_comparisons), rep("OldVSall", l_genes_OldVSall_comparisons) )) ))

pvalues_fdr <- matrix(pvalues_fdr, nrow=1, dimnames=(list("1", c(rep("genes_batch",l_genes_batch_comparisons), rep("OldProt", l_genes_OldProt_comparisons), rep("OldVSall", l_genes_OldVSall_comparisons) )) ))

qvalues_mat <- matrix(qobj$qvalues, nrow=1, dimnames=(list("1", c(rep("genes_batch",l_genes_batch_comparisons), rep("OldProt", l_genes_OldProt_comparisons), rep("OldVSall", l_genes_OldVSall_comparisons) )) ))

sign_frog <- rbind(labels(pvalues_mat[,sign_p_values_index]), all_labels[sign_p_values_index],pvalues_mat[,sign_p_values_index],  pvalues_bonf[sign_p_values_index], pvalues_fdr[sign_p_values_index], qobj$qvalues[sign_p_values_index],  all_common[,sign_p_values_index])
dimnames(sign_frog) <- list(c("Gene list", "Comparison", "pvalue", "Bonferroni correction", "BH correction", "q values", "common genes", rep("", 70)), c(1:length(sign_p_values_index)))


write.table(sign_frog, "sign_frog.csv", sep=",", na="")

#Common genes with an fdr<0.05
sign_fdr_index <-  which(pvalues<0.05 & pvalues_fdr<0.05)
sign_common <- all_common[,sign_fdr_index]

sign_common_mat <- rbind( labels(pvalues_mat[,sign_fdr_index]), all_labels[sign_fdr_index], sign_common
)
dimnames(sign_common_mat) <- list(c("Gene list", "Comparison", "common genes", rep("", 70)), c(1:length(sign_fdr_index)))
#sign_fdr_genes <- unique(sign_common[,!is.na(sign_common) & !sign_common==""])
#sign_fdr_genes

write.table(sign_common_mat, "sign_fdr_genes.csv", sep=",", na="")
```

### Functional annotation:
```{r, echo=FALSE}

library(gProfileR)

result_mm <- apply(sign_common, 2, gprofiler, organism = "mmusculus", significant = T, src_filter = NULL )

result_hs <- apply(sign_common, 2, gprofiler, organism = "hsapiens", significant = T, src_filter = NULL )
```
#a function to combine annotation data with common genes lists to a matrix
fun_annot_mat <- function(mat1, mat2){
  ###mat1 is a matrix containing common genes
  ###mat2 is the output matrix of gprofiler function
  ###output: a matrix containing a list of common genes leading 
  ###two columns with intersection  and functional annotation of these genes
  l_common = dim(mat1)[1]
  l_col = dim(mat1)[2]
  fun_matrix = matrix (NA, nrow = l_common, ncol = l_col*3 )
  
  i=0
  for (k in 1:l_common){
    gprofiler_mat = as.matrix(mat2[k][[1]][, c("intersection", "term.name")])
    gprofiler_mat = apply(gprofiler_mat, 2, function(x){
        length (x) <- l_common; 
        return (x)} )
    fun_matrix[,k+i] = mat1[,k]
    fun_matrix[,k+i+1] = gprofiler_mat[,1]
    fun_matrix[,k+i+2] = gprofiler_mat[,2]
    i = i+2
  }
  fun_matrix
  
}

dim(fun_annot_mat_mm)
fun_annot_mat_mm <- fun_annot_mat(sign_common_mat, result_mm)
n_rep = dim(fun_annot_mat_mm)[2]/3
colnames(fun_annot_mat_mm) <- rep(c("comparison and common genes","functional annotation: genes", "functional annotation: GO terms" ), n_rep)


fun_annot_mat_hs <- fun_annot_mat(sign_common_mat, result_hs)
colnames(fun_annot_mat_hs) <- rep(c("comparison and common genes","functional annotation: genes", "functional annotation: GO terms" ), n_rep)

write.table(fun_annot_mat_mm[,1:771], "funct_annot_sign_fdr_genes_mmusculus_part1.csv", sep=",",  na="")
write.table(fun_annot_mat_mm[,772:1543], "funct_annot_sign_fdr_genes_mmusculus_part2.csv", sep=",",  na="")
write.table(fun_annot_mat_mm[,1544:2316], "funct_annot_sign_fdr_genes_mmusculus_part3.csv", sep=",",  na="")


write.table(fun_annot_mat_hs, "funct_annot_sign_fdr_genes_hsapiens.csv", sep=",",  na="")


```{r,echo=FALSE}

isws_etsi = function(mat1,mat2){

###mat1 is a matrix with the results of functional annotation returned from gprofiler
###mat2 is the matrix containing common genes in each column from each comparison

list_df = list()

for (x in 1:length(mat1)){
dim1 = dim(mat1[x][[1]]["p.value"])[1]

if (dim1==0){
dim1=1
#initialize a matrix
m = matrix(NA, nrow = dim1, ncol = 5)

m[,1] = labels(pvalues_mat[, sign_fdr_index])[x]
m[,2] = all_labels[sign_fdr_index][x]
m[,3] = pvalues[sign_fdr_index][x]
m[,4] = pvalues_fdr[sign_fdr_index][x]
m[,5] = qobj$qvalues[sign_fdr_index][x]
#m[,6] = mat1[x][[1]]["p.value"]
#m[,7] = mat1[x][[1]]["intersection"]
#m[,8] = mat1[x][[1]]["term.name"]

m_df = data.frame(m)

m_df$p.value = 0
m_df$intersection = 0
m_df$term.name = 0
m_df$common_genes = rep(list(mat2[,x][!is.na(mat2[,x]) & !mat2[,x]==""]), dim1)
list_df[[x]] = m_df
}
else {
#initialize a matrix
m = matrix(NA, nrow = dim1, ncol = 5)

m[,1] = labels(pvalues_mat[, sign_fdr_index])[x]
m[,2] = all_labels[sign_fdr_index][x]
m[,3] = pvalues[sign_fdr_index][x]
m[,4] = pvalues_fdr[sign_fdr_index][x]
m[,5] = qobj$qvalues[sign_fdr_index][x]
#m[,6] = mat1[x][[1]]["p.value"]
#m[,7] = mat1[x][[1]]["intersection"]
#m[,8] = mat1[x][[1]]["term.name"]

m_df = data.frame(m)

m_df$p.value = as.matrix(mat1[x][[1]]["p.value"])
m_df$intersection = as.matrix(mat1[x][[1]]["intersection"])
m_df$term.name = as.matrix(mat1[x][[1]]["term.name"])
m_df$common_genes = rep(list(mat2[,x][!is.na(mat2[,x]) & !mat2[,x]==""]), dim1)
list_df[[x]] = m_df
}
}
list_df
}

funct_annot_mm = isws_etsi(result_mm, sign_common)
super_mat_mm = do.call(rbind, funct_annot_mm)
dim(super_mat_mm)
colnames(super_mat_mm) = c("dataset", "comparison", "comparison_pvalue", "comparison_fdr", "comparison_qvalue", "gprofiler_pvalue", "gprofiler_genes", "gprofiler_GO_terms", "common_genes" )

write.table(super_mat_mm[,1:8], "isws_etsi_mus_musculus.csv", sep=",",  na="")

###Homo sapiens
funct_annot_hs = isws_etsi(result_hs, sign_common)

super_mat_hs = do.call(rbind, funct_annot_hs)
dim(super_mat_hs)
colnames(super_mat_hs) = c("dataset", "comparison", "comparison_pvalue", "comparison_fdr", "comparison_qvalue", "gprofiler_pvalue", "gprofiler_genes", "gprofiler_GO_terms", "common_genes" )

write.table(super_mat_hs[,1:8], "isws_etsi_homo_sapiens.csv", sep=",",  na="")
```